---
title: "Microsimulation Project: Formations by Competitions"
author: "Mahdi Munshi"
date: "2024-06-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


While we had a nice attempt at making a microsimulation model before, it had a missive limitation of not considering a lot of important factors; and one of them is overly generalizing across different leagues. In this attempt, I will try to look at the top 5 European leagues, and make the model taking account of that.


## Loading necessary packages and the dataset

```{r}
library(tidyverse)

# Load the dataset
data <- read.csv("Microsimulation_Games_2.csv")

# Check for missing values in the new column as well
data <- data %>%
  drop_na(home_club_formation, away_club_formation, competition_name) %>%
  mutate(winner = ifelse(winner_club_name == home_club_name, "home",
                         ifelse(winner_club_name == away_club_name, "away", "draw")))

```


## Visualizing Match Outcomes by Formations and Competitions

Like the previous attempt, we make another faceted plot visualizing the match outcomes by formation. However, this time each facet representing a different competition (LaLiga, Ligue 1, Serie A, Premier League, and Bundesliga). This allowed us to compare the match outcome distributions across different formation combinations and competitions.

```{r}
library(ggplot2)

# Visualize the distribution of match outcomes by formations and competitions
ggplot(data, aes(x = home_club_formation, fill = winner)) +
  geom_bar(position = "fill") +
  facet_grid(competition_name ~ away_club_formation) +
  labs(title = "Outcome by Formations and Competitions", x = "Home Club Formation", y = "Proportion") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

As usual, it looks very confusing in the knitted report, however, the enlarged version gives decent insights on the relationships between different formations and competitions.


## Training the Predictive Model


Just like before, we are using the caret package to train a multinomial logistic regression model - the exact same model, however, the name of the competition being an important predictor this time.


```{r}
library(caret)


# Train the model
model <- train(winner ~ home_club_formation + away_club_formation + competition_name, data = data,
               method = "multinom")

```


## Simulating Match Outcomes

This time, we changed the function in such way that we can simulate the likely outcome of a match between any two teams, based solely on their chosen formations and most importantly, the competition they are playing in.

```{r}
simulate_match <- function(home_formation, away_formation, competition) {
  predict(model, newdata = data.frame(home_club_formation = home_formation,
                                      away_club_formation = away_formation,
                                      competition_name = competition), type = "prob")
}
```


Let's look at an example of the simulation.

```{r}
simulate_match("3-1-4-2", "4-2-3-1", "LaLiga") *100
```

We simulated a match between two teams playing in the "3-1-4-2" and "4-2-3-1" formations in LaLiga. The simulation showed that the predicted probability of a home win was approximately 41%, the probability of an away win was around 35%, and the probability of a draw was roughly 24%. This feels like a much better model than the previous one, however, football being a complex game, with a lot of factors involved into match outcomes, not just formations, we definitely can make much much better models addressing a lot more factors. However, I am personally satisfied with the model for now, and thus, I am trying to build a shiny app where one can input through dropdowns (do not need to run the fucntion codes and type in the inputs manually) in a seperate R script.



